Certified Symbolic Transducer with the Application of String Solver

### 2025-02-05 Paper/main.tex
stage saving

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\usepackage{stmaryrd}
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite-State Automata (FA) and Finite-State Transducers (FT) are extensively utilized in programming languages and software engineering applications. For instance, regular expressions and their variations play a pivotal role in programming languages like JavaScript, Python, and others. Formalizing FAs and FTs in Coq, Isabelle/HOL, and other proof assistants are a popular topics. However, all these formalization are not practical in real-world applications. 
One of the reasons the transition labels are only single characters in the alphabet, for instance $q\xrightarrow{a} q'$ is a transition and $a$ is a single character. For real-world applications, the alphabet an FA or FT may be enormous or even \emph{infinite}. This classic way of transition definition can yield transitions explosion.

A more practical way is to formalize symoblic FAs [1] and FTs [2], in which transition labels are symbolic and can be infinite. The work of [3] has done the work for FAs, but FTs 
have not yet been formalized, which pose more challenges on proving the correctness of the formalization.

@@

\section{The Key Operation of NFTs}

@@

Figure \ref{fig-def-FAs} shows the definitions of NFA and $\varepsilon-$NFA.
The record type \texttt{eNFA} corresponds to $\varepsilon-$NFA. $\Delta_e$ is same as $\Delta$ in \texttt{NFA}. $\varepsilon$-transitions are defined by $\Delta_e'$.  Now we are ready to present the operation \emph{APPLICATION}. Let $\mathcal{T}$ be an SFT and  $\mathcal{A}$ be an NFA. The output language generated by $\mathcal{T}$ with the input language $\mathcal{A}$ is an an \texttt{eNFA} $\mathcal{A}'$.

@@

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an NFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are intuitive. But we need to explain the role of \texttt{F}. The output function is of type $'a \Rightarrow 'b \text{set option}$, which applies to a single element of the type $'a$. $\texttt{F}$ extends this function to apply it to a set of elements. More precisely, let $f$ be an output function, the semantics of $\texttt{F}$ is defined as $\texttt{F}~f~A=\bigcup_{a\in A} (\text{if }f~a= \text{Some }S \texttt{ then } S \texttt{ else } \emptyset)$.
The transition relations $\Delta_e$ and $\Delta_e'$ are defined by splitting into two cases, whether the transitions in SFTs are $\varepsilon$-transitions or not.

In order to prove this abstract definition of product operation's correctness, we firstly formalize the outputs of an NFT regarding the input of an NFA. Figure \ref{fig-def-output} has two definitions: (1) \texttt{outputE} extracts the output (second element) from a path. (2) \texttt{outputL} defines the set of all outputs of an NFT $\mathcal{T}$, regarding the input language $\mathcal{A}$.

@@

\section{Algorithm Level Refinement}
We have

>>>>

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{intersections}
\usetikzlibrary{automata}
\usetikzlibrary{positioning}
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite-State Automata (FAs) and Finite-State Transducers (FTs) are extensively used in programming languages and software engineering applications. Regular expressions and their variations, for example, play a crucial role in languages like JavaScript, Python, and others. While formalizing FAs and FTs in proof assistants such as Coq and Isabelle/HOL is a popular topic, these formalizations often lack practicality for real-world applications.

One of the reasons the transition labels are only single characters in the alphabet, for instance $q\xrightarrow{a} q'$ is a transition and $a$ is a single character. For real-world applications, the alphabet an FA or FT may be enormous or even \emph{infinite}. This classic way of transition definition can yield transitions explosion.

A more pragmatic approach involves the formalization of symbolic Finite Automata (FAs) [1] and Finite Transducers (FTs) [2], where transition labels are symbolic and potentially infinite. Although the formalization of symbolic FAs has been comprehensively addressed in the work of [3], the formalization of symbolic FTs remains uncharted. This presents additional challenges in establishing the correctness of such formalizations.

@@

\section{The Key Operation of NFTs}
\label{sec-form-fts}

@@

Figure \ref{fig-def-FAs} shows the definitions of NFA and $\varepsilon-$NFA.
The record type \texttt{eNFA} corresponds to $\varepsilon-$NFA. $\Delta_e$ is same as $\Delta$ in \texttt{NFA}. $\varepsilon$-transitions are defined by $\Delta_e'$.  Now we are ready to present the operation \emph{Product}. Let $\mathcal{T}$ be an SFT and  $\mathcal{A}$ be an NFA. The output language generated by $\mathcal{T}$ with the input language $\mathcal{A}$ is an an \texttt{eNFA} $\mathcal{A}'$.

@@

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an NFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are intuitive. But we need to explain the role of \texttt{F}. The output function is of type "$'a \Rightarrow 'b \text{set option}$", which applies to a single element of the type $'a$. $\texttt{F}$ extends this function to apply it to a set of elements. More precisely, let $f$ be an output function, the semantics of $\texttt{F}$ is defined as $\texttt{F}~f~A=\bigcup_{a\in A} (\text{if }f~a= \text{Some }S \texttt{ then } S \texttt{ else } \emptyset)$.
The transition relations $\Delta_e$ and $\Delta_e'$ are defined by splitting into two cases, whether the transitions in SFTs are $\varepsilon$-transitions or not.

$\Delta_e$ has two cases:(1) when $(p, (\text{None}, f),  q)$ is a transition in the NFT. Because the input can be empty ($\varepsilon$), therefore the NFA does not need to move. Therefore the product transtion just output of $f~\text{None}$. (2) when $(p, (Some~\sigma_1, f), q)$ is a transition in the NFT. This transition can only synchronize with the transitions in the NFA that have shared characters with $\sigma_1$. And the output is $F~f~(\sigma_1\cap\sigma_2)$, where $\sigma_2$ is the input label of the NFA. Similarly, $\Delta_e'$ has two cases too.

In order to prove this abstract definition of product operation's correctness, we firstly formalize the outputs of an NFT regarding the input of an NFA. Figure \ref{fig-def-output} has two definitions: (1) \texttt{outputE} extracts the output (second element) from a path. (2) \texttt{outputL} defines the set of all outputs of an NFT $\mathcal{T}$, when the input is in the  language $\mathcal{A}$.

@@

\section{Algorithm Level Refinement}
\label{sec_alg_refinement}
In Section \ref{sec-form-fts}, we give an abstract definition of NFT's product operation. In this section, we present the possible way to compute the Product from programming language's perspective.
In this level, we will also refine the transition labels from \emph{sets} to interval algebra.

\subsection{Intervals}
An interval is defined as a pair $[i, j]$, which corresponds to the set $\{e \mid e \geq i \land e \leq j\}$. In our formalization, we extends the interval definition to a list, that is $[(i_1, j_1), \ldots, (i_n, j_n)]$, which corresponds to the set $\bigcup_{1\leq k\leq n}\{e \mid i_k \leq e \leq i_n\}$. This definition enables us to merge more transitions in NFAs. In the sequent, intervals always mean interval lists.
There are a collection interfaces for intervals. 
\begin{enumerate}
  \item $\texttt{semI}~i$, formalize the semantics of an interval as a set.
  \item $\texttt{empty\_intervals}~i$, checks the emptiness of an interval.
  \item $\texttt{intersectIs}~i_1~i_2$, computes the intersection of two intervals.
  \item $\texttt{diffIs}~i_1~i_2$, computes the intervals difference, that is the interval corresponds to the set $\texttt{semI}(i_1)-\texttt{semI}(i_2)$.
\end{enumerate}

\subsection{Algorithm of Product Computation.}

Fig. \ref{fig-compute-nft-product} illustrates the computation of NFT products. \texttt{prods\_imp} computes the product of two sets. It is illustrated in \ref{fig-def-prods_imp}. We use the constructs from \emph{Refine\_Monadic}. The algorithm itself is intuitive: the nested two loops traverse all possible combination of states in the states.

\begin{figure}[ht]
	\begin{lstlisting}
definition productT_impl where
  "productT_impl |$\mathcal{T}$| |$\mathcal{A}$| F fe = do {
    Q |$\leftarrow$| prods_imp (nft_states |$\mathcal{T}$|) (nfa_states |$\mathcal{A}$|);
    (D1, D2) |$\leftarrow$| trans_comp_imp (nft_tranfun |$\mathcal{T}$|) F fe 
        (nft_trans |$\mathcal{T}$|) (nfa_trans |$\mathcal{A}$|) (nfa_states |$\mathcal{A}$|);
    I |$\leftarrow$| prods_imp (nft_initial |$\mathcal{T}$|) (nfa_initial |$\mathcal{A}$|);
    F |$\leftarrow$| prods_imp (nft_accepting |$\mathcal{T}$|) (nfa_accepting |$\mathcal{A}$|);
    RETURN (Q,D1,D2,I, F)
  }"
\end{lstlisting}
\caption{The computation of NFT product}
\label{fig-compute-nft-product}
\end{figure}

\begin{figure}[ht]
	\begin{lstlisting}
definition prods_imp where
  "prods_imp Q1 Q2 =
   FOREACH {q. q |$\in$| s.|$\alpha$| Q1} (|$\lambda$| q Q. do {
    S |$\leftarrow$| FOREACH {q. q |$\in$| s.|$\alpha$| Q2}
         (|$\lambda$| q' Q'. RETURN (ss.ins (q,q') Q')) (ss.empty ());
    RETURN (ss.union Q S)
   }) (ss.empty ())"
\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-prods_imp}
\end{figure}

The challenging part is how to compute the transitions sets. It is illustrated in Fig. \ref{fig-def-prods-imp}. The function \texttt{trans\_comp\_imp} traverse the transtions in the NFT and split each transtion into two cases. This function uses another two functions:

1. \texttt{copy\_tran\_impl}

2. \texttt{copy\_tran'\_impl}

\begin{figure}[ht]
	\begin{lstlisting}
definition trans_comp_imp where
  "trans_comp_imp M F fe T1 T2 Q =
    FOREACH {t. t |$\in$| dt.|$\alpha$| T1}
     (|$\lambda$| (q, (|$\alpha$|, f), q') (D1, D2). 
       (if (|$\alpha$| = None) then 
          (if ((M f) None = None) then 
            do {
              D2' |$\leftarrow$| copy_tran_imp q q' Q;
              RETURN (D1, ssd.union D2 D2')
            }
            else do {
              D1' |$\leftarrow$| copy_tran'_imp q (the (M f None)) q' Q;
              RETURN (ddt.union D1 D1', D2)
            })
          else 
            (subtrans_comp_imp M q (the |$\alpha$|) f q' F fe T2 D1 D2)
        )) (ddt.empty, ssd.empty ())"

\end{lstlisting}
\caption{The computation of transition sets}
\label{fig-def-prods-imp}
\end{figure}

\section{Refinement to Efficient Data Structure}

\section{An Application to String Solver}
\label{sec-app-str-solver}
In this section, we will show how to use the formalized symbolic transducer to solve the string constraints with replacement operations.

A replacement operation has the following form: $\texttt{replace}(\texttt{str}, \texttt{pattern}, \texttt{replacement})$. This operation replaces the occurrence of the substring in \texttt{str} that matches \texttt{pattern} with \texttt{replacement}. \texttt{str} and \texttt{replacement} are strings and \texttt{pattern} is a regular expression.

The semantics of the replacement operation can be described in two cases: (1) if there are strings in \texttt{str} that match \texttt{pattern}, then replace them with \texttt{replacement}. (2) If there are no matches, then the result is \texttt{str}.

The first case can be modeled by the SFTs. We illustrate this case by an example. Given a replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$, we can model this operation by an SFT.

There are three steps to model this operation:

\begin{enumerate}
  \item Generate an SNFA from the regular expression "$[0-9]+$". Adding output functions to the transitions of the SNFA that outputs empty characters to make it an SFT.
  \item Generate an SNFA from the replacement string "$"$\text{ITP}$"$". Note that a single string is also a regular expression. Adding input labels and output functions to the transitions of the SNFA to make it an SFT.
  \item Concatenate the two SFTs and adding preceding and succeeding transitions to make it accept the whole input string.
\end{enumerate}

We explain it step by step.

\noindent\emph{Step 1.}
The SNFA corresponding to the regular expression "$[0-9]+$" is shown in Figure \ref{fig-snfa-pattern}. On the left side is the SNFA and on the right side is the corresponding SFT. The function \texttt{f} is the output function that outputs empty characters. It can be defined as $\texttt{f} = \lambda x.~\texttt{None}$. The function \texttt{g} is the output function that outputs the characters in the replacement string. It can be defined as (the ascii code of $'I'$, $'T'$, $'P'$ are 73, 84, 80 respectively):

\[
\texttt{g} = \lambda~i~x.~\texttt{match }~i~ with~ 1 \rightarrow [(73, 73)] ~|~ 2 \rightarrow [(84, 84)] ~|~ 3 \rightarrow [(80, 80)] ~|~ \_ ~\rightarrow \texttt{None}
\]

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state, accepting] (q1) [right=of q0] {$q_1$};

   \path[->] 
   (q0) edge [bend left] node {[0-9]} (q1)
   (q1) edge [loop above] node {[0-9]} (q1);

  \node[state, initial] (q0') [right=4cm of q1] {$q_0'$}; 
  \node[state, accepting] (q1') [right=of q0'] {$q_1'$};

   \path[->] 
   (q0') edge [bend left] node {[0-9]/\texttt{f}} (q1')
   (q1') edge [loop above] node {[0-9]/\texttt{f}} (q1');
\end{tikzpicture}
\caption{The SFA for "$[0-9]+$" duplicated}
\label{fig-snfa-pattern}
\end{figure}

\noindent\emph{Step 2.}
The SNFA corresponding to the replacement string "\text{ITP}" is shown in Figure \ref{fig-snfa-replacement}. On the left side is the SNFA and on the right side is the corresponding SFT. The function \texttt{f} is the output function that outputs the characters in the replacement string. It can be defined as $\texttt{f} = \lambda x.~\texttt{Some } x$.

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
  % Original NFA for "ITP"
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state] (q1) [right=of q0] {$q_1$}; 
  \node[state] (q2) [below=of q0] {$q_2$}; 
  \node[state, accepting] (q3) [right=of q2] {$q_3$};

  \path[->] 
  (q0) edge node {I} (q1)
  (q1) edge node {T} (q2)
  (q2) edge node {P} (q3);

  % Duplicate NFA for "ITP" on the right
  \node[state, initial] (q0') [right=4cm of q1] {$q_0'$}; 
  \node[state] (q1') [right=of q0'] {$q_1'$}; 
  \node[state] (q2') [below=of q0'] {$q_2'$}; 
  \node[state, accepting] (q3') [right=of q2'] {$q_3'$};

  \path[->] 
  (q0') edge node {I/\texttt{g 1}} (q1')
  (q1') edge node {T/\texttt{g 2}} (q2')
  (q2') edge node {P/\texttt{g 3}} (q3');
\end{tikzpicture}
\caption{The SFA for "$"\text{ITP}"$" duplicated}
\label{fig-snfa-replacement}
\end{figure}

\noindent\emph{Step 3.}
The concatenation of the two SFTs is shown in Figure \ref{fig-rearranged-automata}. It contains two steps: (1) rearrange the transitions of the two SFTs to make them accept the whole input string. (2) add the preceding and succeeding transitions to make it accept the whole input string.

\begin{figure}[h] \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
    % Second automaton from the first figure
    \node[state, initial] (q0') {$q_0'$}; 
    \node[state] (q1') [right=of q0'] {$q_1'$};

     \path[->] 
     (q0') edge [bend left] node {[0-9]/\texttt{f}} (q1')
     (q1') edge [loop above] node {[0-9]/\texttt{f}} (q1')
     (q0') edge [loop below] node {[\texttt{None}/\texttt{f}]} (q0');

    % Second automaton from the second figure
    \node[state] (q0'') [right=4cm of q1'] {$q_0''$}; 
    \node[state] (q1'') [right=of q0''] {$q_1''$}; 
    \node[state] (q2'') [below=of q0''] {$q_2''$}; 
    \node[state, accepting] (q3'') [right=of q2''] {$q_3''$};

    \path[->] 
    (q0'') edge node {I/\texttt{g 1}} (q1'')
    (q1'') edge node {T/\texttt{g 2}} (q2'')
    (q2'') edge node {P/\texttt{g 3}} (q3'')
    (q3'') edge [loop below] node {[\texttt{None}/\texttt{f}]} (q3'')
    (q1') edge node {[\texttt{None}/\texttt{f}]} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$}
  \label{fig-rearranged-automata}
  \end{figure}
